The bhsdtr package: a general purpose method of bayesian inference for Signal Detection Theory models

### 2018-02-12 inst/preprint/paper.tex
paper added

In what follows, after introducing the most common version of the SDT
model, we describe its generalization which can accomodate data from
rating experiments. Next, we explain briefly why, if a method of
inference for SDT models were to be of general use in psychology
studies, it is essential that it was based on a model equipped with
the general hierarchical linear regression structure. The
\texttt{bhsdtr} package meets this requirement thanks to a novel
parametrization. We describe this novel parametrization in great
detail and we motivate it by explaining how the reliance on a
relatively standard parametrisation leads to problems in the two other
available implementations. We end the first part of this paper with a
formal definition of the model as implemented in \texttt{bhsdtr}. The
second part contains an overview of the package and a tutorial in
which we demonstrate how to use our method in practice. Before we go
any further, however, a note on terminology seems in order.

@@

In the context of hierarchical modelling factors such as participants,
items, or replications are often referred to as groups. In our opinion
this naming convention may be confusing; a single participant is both
a group and a member of some group, while at the same time the term
"group" is perhaps most strongly associated with study conditions, as
in "experimental group". In this paper we use the term "sampled
factor" instead, because, by virtue of being a new term, it is
unambiguous, and because it seems descriprively correct: the term
"sampled factor" captures all the essential properties of such
variables, i.e., a nominal scale, the fact that values are sampled
from a larger population and are usually not of direct interest, as in
"this is only a sample", and that conclusions
% ?? jeżeli już mamy jakąś analizę statystyczną, to zakładamy, w
% uproszczeniu, że target population jest określona
of statistical analysis are meant to apply to the whole population of
possible values.

@@

According to Signal Detection Theory each stimulus in a classification
task gives rise, by some unspecified cognitive process, to an
unidimensional internal evidence value $s$ sampled from
% ?? stimulus już był wprowadzony, więc jego klasa jest (jakaś, ale)
% określona
a distribution that depends on the stimulus class. For historical
reasons the two stimulus classess are often referred to as "noise" and
"signal", and task performance is described in terms of hits, correct
rejections, omissions, and false alarms, but this terminology is
appropriate only when the model is applied to tasks that require
detection, which is by far not always the case. In the most widely
used version of the model, shown in Fig.~\ref{basicsdt}, the two
evidence distributions are normal with the same variance, which is
usually fixed at unity to make the model identifiable. The distance
$d'$ between the means of the
% ?? sensitivity bez przedimka?
evidence distributions represents sensitivity. Because normal
distributions are unbounded, $s$ is always ambiguous, and so a
criterion $c$ placed on the evidence axis has to be used to reach a
discrete
% ?? bias bez przedimka?
decision. A participant is assumed to decide that a stimulus belongs
to the first class (e.g., an old item) if $s < c$, or that it belongs
to the second class (e.g., a new item) if $s \geq c$. The location of
the decision criterion $c$ represents bias.

@@

When SDT models are used in psychology studies, researchers are
usually interested in relationships between SDT parameters and
additional measured or manipulated
% ?? the dependence, bo ogólnie rozumiana kwestia zależności, stimulus
% strength - ogólne zjawisko, a nie jakaś konkretna, albo pojedyncza,
% ale bliżej nieokreślona
variables; A good example is the dependence of $d'$ on stimulus
strength. Also, in a great majority of psychology studies in which
classification tasks are used the data have hierarchical structure,
i.e., there are repeated measures for each participant or item and
participants or items are only samples from the target population. It
seems worthwhile to explain why a general-purpose method of inference
for SDT models should accomodate both kinds of situations.

@@

% mówimy o ogólnym co prawda, ale zarazem pojedynczym/unikalnym
% problemie: the separation? Chętnie bym się pozbył przedimka
Another common source of problems is the separation of estimation of
non-linear model parameters from regression analysis whichs aim to
relate these parameters to measured or manipulated variables. When the
SDT parameters are estimated separately for each subject and
condition, and only later regressed on predictors of interest, a
number of additional issues may arise.

@@

The Gibbs sampler created by \citeA{morey2008problematic} allows for
at most two sampled factors to have independent normally distributed
random effects on the evidence distribution means. Unlike $d'$, each
evidence distribution mean considered in isolation is an unconstrained
parameter, but the mean of the second one is by definition greater
($d' > 0$), or equal ($d' = 0$), than the mean of the first one. The
authors explicitly admit that their algorithm does not enforce this
restriction, because, as they claim in the paper, it would greatly
complicate analysis. At the same time the authors fail to mention that
an immediate consequence of allowing for negative $d'$ values is that
the resulting posterior distribution looses its intended
interpretation, since it can have a non-zero mass when $d' < 0$. The
outermost criteria are fixed at $0$ and $1$, and the oredering
restriction is enforced by assuming that the likelihood is $0$
whenever $\bm{c}_{i+1} \leq \bm{c}_i$. As the authors explain, because
a sampled factor can have independent random effects on the evidence
distribution means, it can have an effect on all the criteria:
shifting both means by the same amount in the same direction is
equivalent to keeping the sensitivity intact, while shifting the
criteria relative to the evidence distributions. However, the elements
of the criteria vector cannot be affected differently by the same
sampled factor, which is an unrealistic restriction.

@@

\noindent where $\Phi$ is the CDF of the standard normal distrubution
and $\bm{\gamma} \in R^K$, with $\bm{\gamma}_K$ fixed at $0$ for
identifiability. The idea is illustrated in Fig.~\ref{fig:2} below:

@@

Once $d'$ and $c$ are derived from the unconstrained $\delta$ and
$\gamma$ parameters, the SDT model can be supplemented with a
hierarchical linear regression structure. To avoid having to deal with
an even more complicated index notation\footnote{The reader familiar
  with hierarchical models may be surprised by our use of superscrip
  parenthesised greek letters to express hierarchical
  relationships. We chose this convention because it allowed us to use
  subscripts to denote elements of vectors and matrices while
  minimizing the number of nested sub- or superscripts.}, below we
have presented only the simple case of one sampled factor.

@@

This function requires at least three agruments; a stanfit object, an
aggregated data list produced by the \texttt{aggregate\_responses}
function and that was used to produce the stanfit object, and a vector
of names of variables that will determine how the data will be
partitioned before plotting. We recommend assessing the fit at the
individual level, but we did not include the participant
identification number in the list of conditioning variables, because
the resulting plot would take up too much space.

>>>>

In what follows, after introducing the most common version of the SDT
model, we describe its generalization which can accommodate data from
rating experiments. Next, we explain briefly why, if a method of
inference for SDT models were to be of general use in psychology
studies, it is essential that it was based on a model equipped with
the general hierarchical linear regression structure. The
\texttt{bhsdtr} package meets this requirement thanks to a novel
parametrization. We describe this novel parametrization in great
detail and we motivate it by explaining how the reliance on a
relatively standard parametrisation leads to problems in the two other
available implementations. We end the first part of this paper with a
formal definition of the model as implemented in \texttt{bhsdtr}. The
second part contains an overview of the package and a tutorial in
which we demonstrate how to use our method in practice. Before we go
any further, however, a note on terminology seems in order.

@@

In the context of hierarchical modelling factors such as participants,
items, or replications are often referred to as groups. In our opinion
this naming convention may be confusing; a single participant is both
a group and a member of some group, while at the same time the term
"group" is perhaps most strongly associated with study conditions, as
in "experimental group". In this paper we use the term "sampled
factor" instead, because, by virtue of being a new term, it is
unambiguous, and because it seems descriptively correct: the term
"sampled factor" captures all the essential properties of such
variables, i.e., a nominal scale, the fact that values are sampled
from a larger population and are usually not of direct interest, as in
"this is only a sample", and that conclusions
% ?? jeżeli już mamy jakąś analizę statystyczną, to zakładamy, w
% uproszczeniu, że target population jest określona
of statistical analysis are meant to apply to the whole population of
possible values.

@@

According to Signal Detection Theory each stimulus in a classification
task gives rise, by some unspecified cognitive process, to an
unidimensional internal evidence value $s$ sampled from
% ?? stimulus już był wprowadzony, więc jego klasa jest (jakaś, ale)
% określona
a distribution that depends on the stimulus class. For historical
reasons the two stimulus classes are often referred to as "noise" and
"signal", and task performance is described in terms of hits, correct
rejections, omissions, and false alarms, but this terminology is
appropriate only when the model is applied to tasks that require
detection, which is by far not always the case. In the most widely
used version of the model, shown in Fig.~\ref{basicsdt}, the two
evidence distributions are normal with the same variance, which is
usually fixed at unity to make the model identifiable. The distance
$d'$ between the means of the
% ?? sensitivity bez przedimka?
evidence distributions represents sensitivity. Because normal
distributions are unbounded, $s$ is always ambiguous, and so a
criterion $c$ placed on the evidence axis has to be used to reach a
discrete
% ?? bias bez przedimka?
decision. A participant is assumed to decide that a stimulus belongs
to the first class (e.g., an old item) if $s < c$, or that it belongs
to the second class (e.g., a new item) if $s \geq c$. The location of
the decision criterion $c$ represents bias.

@@

When SDT models are used in psychology studies, researchers are
usually interested in relationships between SDT parameters and
additional measured or manipulated
% ?? the dependence, bo ogólnie rozumiana kwestia zależności, stimulus
% strength - ogólne zjawisko, a nie jakaś konkretna, albo pojedyncza,
% ale bliżej nieokreślona
variables; A good example is the dependence of $d'$ on stimulus
strength. Also, in a great majority of psychology studies in which
classification tasks are used the data have hierarchical structure,
i.e., there are repeated measures for each participant or item and
participants or items are only samples from the target population. It
seems worthwhile to explain why a general-purpose method of inference
for SDT models should accommodate both kinds of situations.

@@

% mówimy o ogólnym co prawda, ale zarazem pojedynczym/unikalnym
% problemie: the separation? Chętnie bym się pozbył przedimka
Another common source of problems is the separation of estimation of
non-linear model parameters from regression analysis which aim to
relate these parameters to measured or manipulated variables. When the
SDT parameters are estimated separately for each subject and
condition, and only later regressed on predictors of interest, a
number of additional issues may arise.

@@

The Gibbs sampler created by \citeA{morey2008problematic} allows for
at most two sampled factors to have independent normally distributed
random effects on the evidence distribution means. Unlike $d'$, each
evidence distribution mean considered in isolation is an unconstrained
parameter, but the mean of the second one is by definition greater
($d' > 0$), or equal ($d' = 0$), than the mean of the first one. The
authors explicitly admit that their algorithm does not enforce this
restriction, because, as they claim in the paper, it would greatly
complicate analysis. At the same time the authors fail to mention that
an immediate consequence of allowing for negative $d'$ values is that
the resulting posterior distribution looses its intended
interpretation, since it can have a non-zero mass when $d' < 0$. The
outermost criteria are fixed at $0$ and $1$, and the ordering
restriction is enforced by assuming that the likelihood is $0$
whenever $\bm{c}_{i+1} \leq \bm{c}_i$. As the authors explain, because
a sampled factor can have independent random effects on the evidence
distribution means, it can have an effect on all the criteria:
shifting both means by the same amount in the same direction is
equivalent to keeping the sensitivity intact, while shifting the
criteria relative to the evidence distributions. However, the elements
of the criteria vector cannot be affected differently by the same
sampled factor, which is an unrealistic restriction.

@@

\noindent where $\Phi$ is the CDF of the standard normal distribution
and $\bm{\gamma} \in R^K$, with $\bm{\gamma}_K$ fixed at $0$ for
identifiability. The idea is illustrated in Fig.~\ref{fig:2} below:

@@

Once $d'$ and $c$ are derived from the unconstrained $\delta$ and
$\gamma$ parameters, the SDT model can be supplemented with a
hierarchical linear regression structure. To avoid having to deal with
an even more complicated index notation\footnote{The reader familiar
  with hierarchical models may be surprised by our use of superscript
  parenthesised Greek letters to express hierarchical
  relationships. We chose this convention because it allowed us to use
  subscripts to denote elements of vectors and matrices while
  minimizing the number of nested sub- or superscripts.}, below we
have presented only the simple case of one sampled factor.

@@

This function requires at least three arguments; a stanfit object, an
aggregated data list produced by the \texttt{aggregate\_responses}
function and that was used to produce the stanfit object, and a vector
of names of variables that will determine how the data will be
partitioned before plotting. We recommend assessing the fit at the
individual level, but we did not include the participant
identification number in the list of conditioning variables, because
the resulting plot would take up too much space.
